---
title: "Motor Trend Data Analysis"
author: "Jeffrey M. Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    keep_md: no
    df_print: paged
  pdf_document:
    toc: yes
    df_print: kable
    number_sections: false
    fig_caption: yes
    highlight: tango
    dev: pdf
  word_document:
    toc: yes
    df_print: paged
    keep_md: no
---

## Course Project

**Regression Models Course Project**

Peer-graded Assignment

* This course project is available on GitHub

    <a href="https://github.com/oraclejavanet/regression-models-course-project" target="_top">Motor Trend Data Analysis</a>

## Executive Summary

This analysis is being performed for Motor Trend, a popular American automobile
magazine, to evaluate the relationship between transmission type (manual or 
automatic) and fuel consumption in miles per gallon (MPG) in automobiles. The
analysis extends beyond transmission type to also include other possible
variables that explain variance in fuel consumption (MPG).

As part of this analysis, Motor Trend is particularly interested in the
following two questions:

* Is an automatic or manual transmission better for MPG
* Quantify the MPG difference between automatic and manual transmissions

The analysis will be conducted using exploratory data analyses and
linear regression models using the `mtcars` dataset.

## Data Description

The mtcars dataset is comprised of data that was was extracted from the
1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of
automobile design and performance for 32 automobiles (1973-74 models).

The mtcars dataset is a data frame with 32 observations on 11 (numeric)
variables:

* `mpg` Miles/(US) gallon
* `cyl` Number of cylinders
* `disp` Displacement (cu.in.)
* `hp` Gross horsepower
* `drat` Rear axle ratio
* `wt` Weight (1000 lbs)
* `qsec` 1/4 mile time
* `vs` Engine (0 = V-shaped, 1 = straight)
* `am` Transmission (0 = automatic, 1 = manual)
* `gear` Number of forward gears
* `carb` Number of carburetors

## Environment Setup

```{r load-packages, message = FALSE, echo = TRUE}
if (!require(knitr)) {
    install.packages("knitr")
    library(knitr)
}
if (!require(kableExtra)) {
    install.packages("kableExtra")
    library(kableExtra)
}
if (!require(ggplot2)) {
    install.packages("ggplot2")
    library(ggplot2)
}
if (!require(GGally)) {
    install.packages("GGally")
    library(GGally)
}
if (!require(MASS)) {
    install.packages("MASS")
    library(MASS)
}
```

```{r setup, include = FALSE}
# set knitr options
knitr::opts_chunk$set(echo = TRUE, fig.path = 'figures/')

# clear global workspace (including hidden objects) for this session
rm(list = ls(all.names = TRUE))

# free up memory and display statistics on free memory
gc()

# disable scientific notation for numbers
options(scipen = 1)

# knit hook to allow partial output from a code chunk
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options)) # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines) == 1) { # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
```

## Load Data

Load the mtcars dataset and display the internal structure of the variables.

```{r load-data, echo = TRUE}
library(datasets)
data(mtcars)
str(mtcars)
```

## Data Analysis

Perform some basic exploratory and inferential data analysis of the data to
study the relationship between transmission type (manual or automatic) and
automobile fuel consumption in miles per gallon (MPG).

### Basic Data Summary

The mtcars dataset includes 1 target variable (mpg) and 10 independent control
variables with 32 observations.

See [A.1 Basic Data Summary](#a.1-basic-data-summary) in the Appendix section
which shows the range and quartiles for each variable.

### Relative Mean

Display the relative mean of automobile fuel consumption data grouped by
transmission type.

```{r exploratory-data-analysis-relative-mean, echo = TRUE}
by(data = mtcars$mpg,
   INDICES = list(factor(mtcars$am, labels = c("Automatic", "Manual"))),
   summary)
```

### Impact of Transmission Type on Fuel Consumption

[Figure A.2.1 (Boxplot)](#figure-a.2.1-boxplot) and
[Figure A.2.2 (Histogram)](#figure-a.2.2-histogram) in the Appendix section
plot the relationship between transmission type and fuel consumption in
automobiles.

### Inferential Statistics

Hypothesis testing will be conducted to study the impact of transmission type on
automobile fuel consumption. A t-test will be performed on the null
hypothesis that transmission type has no effect on automobile fuel consumption.

```{r inferential-statistics-test-1, echo = TRUE}
t1 <- t.test(mpg ~ am, data = mtcars, conf.level = 0.95)
paste0("p-value = ", round(t1$p.value, 4))
paste0("confidence interval = (",
       round(t1$conf.int[1], 4),
       ", ",
       round(t1$conf.int[2], 4), ")")
t1$estimate
```

The observed p-value `r round(t1$p.value, 4)` is less than 0.05 and the 95%
confidence interval does not contain zero. This indicates strong evidence
against the null hypothesis so the null hypothesis can be rejected.

### Observation

The calculated mean for both transmission types and the
[provided plots](#a.2-plot-impact-of-transmission-type-on-fuel-consumption)
show a
significant increase in better fuel consumption for automobiles with a manual
transmission versus automatic.

The difference between mean fuel consumption of automatic and manual
transmission is significantly different where the estimated difference favors
a manual transmission by
`r round(t1$estimate[2] - t1$estimate[1], 2)` MPG.

## Linear Models

Linear regression analysis will be used to extend beyond our initial interest
in the relationship between the transmission type variable only and fuel
consumption. Other possible variables in the mtcars dataset may better explain
variance in fuel consumption.

### Simple Linear Regression Model

Based on our initial interest with only transmission type, start by building a
simple linear regression model between the response variable (MPG) and the
single predictor (transmission type).

```{r simple-linear-regression-model, echo = TRUE}
singleModelFit <- lm(mpg ~ am, data = mtcars)
summary(singleModelFit)
```

Here we see that the adjusted $R^2$ is only 0.3385 which suggests that this
model can only explain 33.8% of the variance in fuel consumption (approximately
one third) based on transmission type alone.

Looking to the correlation table below, it's possible that other variables in
the dataset can better explain the outcome:

```{r correlation-table, echo = TRUE}
round(cor(mtcars, method = "pearson")[1,], 2)
```

Also see [A.3 Correlation Matrix Plot](#a.3-correlation-matrix-plot) in the
Appendix section which provides a plot showing the correlation coefficients
between all variables.

### Multiple Linear Regression Model

Perform stepwise regression using the `stepAIC()` function from the `MASS`
package to find the subset of variables which result in the best model fit (a
model that lowers prediction error). 

Start by building an initial model with all variables as predictors. Stepwise
regression will select the significant predictors for the final model which is
the best model. The AIC algorithm runs `lm` multiple times to build multiple
regression models and selects the best variables from them using both forward
selection and backward elimination methods.

```{r stepwise-regression-code, echo = TRUE, results = "hide"}
initialModel <- lm(mpg ~ ., data = mtcars)
stepReg <- stepAIC(initialModel, direction = "both")
```

See [A.4 Stepwise Regression](#a.4-stepwise-regression) for stepwise regression
output.

Show results of stepwise regression variable selection.

```{r stepwise-regression-results, echo = TRUE, output.lines=1:8}
print(stepReg$anova)
```

As shown above, the best model obtained from the stepwise regression procedure
consists of the predictors weight (wt) and 1/4 mile time (qsec) in addition to
transmission type (am).

```{r best-model-fit, echo = TRUE}
bestModelFit <- lm(mpg ~ wt + qsec + am, data = mtcars)
summary(bestModelFit)
```

Here we observe that the adjusted $R^2$ is now 0.8336 which suggests that the
new model (including the three predictors) can explain 84% of the variance in
fuel consumption.

## Analysis of Residuals

*Pending*

## Conclusion

This analysis concludes the following:

1. **Is an automatic or manual transmission better for MPG?**

    Automobiles with a manual transmission yield better gas mileage than
    vehicles with an automatic transmission. However, determining fuel
    consumption based on transmission type alone showed that the relationship
    was not as statistically significant as first thought. Models were built
    with confounding variables such as weight (wt) and 1/4 mile time (qsec)
    in addition to transmission type (am) that better explain variance in fuel
    consumption.

1. **Quantify the MPG difference between automatic and manual transmissions.**

    Based on our simple linear regression model that only considered
    transmission type, the mean difference in fuel consumption increased to
    `r round(coefficients(singleModelFit)[Intercept = "am"], 2)` MPG favoring a
    manual transmission.
    
    However, when the variables weight (wt) and 1/4 mile time (qsec) were added
    to the best fitted multiple regression model, the advantage of the
    manual transmission decreased to
    `r round(coefficients(bestModelFit)[Intercept = "am"], 2)`.
    
## Appendix

### A.1 Basic Data Summary

Provide a basic summary of the data.

```{r appendix-basic-data-summary, echo = TRUE, results = 'hold'}
# target variable
kable(summary(mtcars[1]),
              row.names = FALSE,
              align = c("l"),
              caption = "Target Variable (MPG)") %>%
                kable_styling(position = "center")

# independent control variables
kable(summary(mtcars[2:6]),
              row.names = FALSE,
              align = c(rep("l", 5)),
              caption = "Control Variables") %>%
                kable_styling(position = "left")

kable(summary(mtcars[7:11]),
              row.names = FALSE,
              align = rep('l', 5),
              caption = "Control Variables (cont)") %>%
                kable_styling(position = "left")
```

### A.2 Plot Impact of Transmission Type on Fuel Consumption

Plot the relationship of automobile fuel consumption as a function of
transmission type.

#### Figure A.2.1 (Boxplot)

```{r mpg-on-trans-type-bp, echo = TRUE}
g <- ggplot(data = mtcars,
            aes(x = factor(am, labels = c("Automatic", "Manual")),
                y = mpg, fill = factor(am, labels = c("Automatic", "Manual"))))
g <- g + geom_boxplot()
g <- g + scale_colour_discrete(name = "Transmission Type")
g <- g + scale_fill_discrete(name = "Transmission Type")
g <- g + xlab("Transmission Type")
g <- g + ylab("Fuel Consumption (MPG)")
g <- g + theme(plot.title = element_text(size = 14,
                                         hjust = 0.5,
                                         vjust = 0.5,
                                         margin = margin(b = 12, unit = "pt")),
           axis.text.x = element_text(angle = 0,
                                      hjust = 0.5,
                                      vjust = 0.5,
                                      margin = margin(b = 10, unit = "pt")),
           axis.text.y = element_text(angle = 0,
                                      hjust = 0.5,
                                      vjust = 0.5,
                                      margin = margin(l = 10, unit = "pt")))
g <- g + ggtitle("Impact of Transmission Type on Fuel Consumption")
print(g)
```

#### Figure A.2.2 (Histogram)

```{r mpg-on-trans-type-hist, echo = TRUE}
g <- ggplot(data = mtcars, aes(x = mpg, y = ..count..))
g <- g + geom_histogram(binwidth = 0.75,
                        aes(fill = factor(am, labels = c("Automatic", "Manual"))))
g <- g + facet_grid(. ~ factor(am, labels = c("Automatic", "Manual")))
g <- g + scale_colour_discrete(name = "Transmission Type")
g <- g + scale_fill_discrete(name = "Transmission Type")
g <- g + xlab("Fuel Consumption (MPG)")
g <- g + ylab("Frequency")
g <- g + theme(plot.title = element_text(size = 14,
                                         hjust = 0.5,
                                         vjust = 0.5,
                                         margin = margin(b = 12, unit = "pt")),
           axis.text.x = element_text(angle = 0,
                                      hjust = 0.5,
                                      vjust = 0.5,
                                      margin = margin(b = 10, unit = "pt")),
           axis.text.y = element_text(angle = 0,
                                      hjust = 0.5,
                                      vjust = 0.5,
                                      margin = margin(l = 10, unit = "pt")))
g <- g + ggtitle("Impact of Transmission Type on Fuel Consumption")
print(g)
```

### A.3 Correlation Matrix Plot

Show the correlation coefficients for all variables.

```{r correlation-matrix, echo = TRUE}
ggcorr(data = mtcars,
       method = c("pairwise", "pearson"),
       nbreaks=8,
       palette='RdGy',
       label=TRUE,
       label_size=5,
       label_color='white')
```

### A.4 Stepwise Regression

```{r stepwise-regression-appendix, echo = TRUE, results = 'hold'}
initialModel <- lm(mpg ~ ., data = mtcars)
stepReg <- stepAIC(initialModel, direction="both")
```

