---
title: "Motor Trend Data Analysis"
author: "Jeffrey M. Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    keep_md: no
    df_print: paged
  pdf_document:
    toc: yes
    df_print: kable
    number_sections: false
    fig_caption: yes
    highlight: tango
    dev: pdf
  word_document:
    toc: yes
    df_print: paged
    keep_md: no
---

## Course Project

**Regression Models Course Project**

Peer-graded Assignment

* This course project is available on GitHub

    <a href="https://github.com/oraclejavanet/regression-models-course-project" target="_top">Motor Trend Data Analysis</a>

## Synopsis

This analysis is being performed for Motor Trend, a popular American automobile
magazine, to evaluate the relationship between transmission type (manual or 
automatic) and miles per gallon (MPG) in automobiles. The analysis extends
beyond transmission design to also include other possible variables such as
number of cylinders, gross horsepower and displacement that explain variance in
MPG.

Motor Trend is particularly interested in the following two questions:

* Is an automatic or manual transmission better for MPG
* Quantify the MPG difference between automatic and manual transmissions


=============



In this project we take the mtcars data set and write up an analysis to answer their question using linear regression models and exploratory data analyses.

using the `mtcars` dataset.

In this project, we will analyze the mtcars dataset from the 1974 Motor Trend US magazine to answer the following questions:
 
 
 

## Environment Setup

```{r setup, include = FALSE}
# set knitr options
knitr::opts_chunk$set(echo = TRUE, fig.path='figures/')

# clear global workspace (including hidden objects) for this session
rm(list = ls(all.names = TRUE))

# free up memory and display statistics on free memory
gc()

# disable scientific notation for numbers
options(scipen = 1)
```

Load packages used in this analysis.

```{r load-packages, echo = TRUE}
if (!require(ggplot2)) {
    install.packages("ggplot2")
    library(ggplot2)
}
if (!require("dplyr")) {
    install.packages("dplyr")
    library("dplyr")
}
```

Display session information.

```{r display-session-info, echo = TRUE}
sessionInfo()
```

## Load Data

Load and prepare the mtcars dataset. Convert categorical variables to factors.

```{r load-data, echo = TRUE}
library(datasets)
data(mtcars)
```

## Basic Data Summary

After loading the mtcars dataset, provide a basic summary of the data.

```{r basic-data-summary, echo = TRUE}
str(mtcars)
summary(mtcars)
head(mtcars)
dim(mtcars)

# tabulate delivery method and dosage level values
#table(ToothGrowth$supp, ToothGrowth$dose)

# summary of tooth length data grouped by delivery method and dosage level
#by(data = ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), summary)
```

## Exploratory Data Analysis

Perform some basic exploratory data analyses of the data. The analyses will
explore the following relations:

1. Relation 1
1. Relation 2

### Relation 1

```{r relation-1, echo = TRUE}
```

**Observation**

*Pending*

### Relation 2

```{r relation-2, echo = TRUE}
```

**Observation**

*Pending*



## Linear Models


*Pending*


## Analysis of Residuals


*Pending*


## Conclusion


*Pending*



