---
title: "Motor Trend Data Analysis"
author: "Jeffrey M. Hunter"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    keep_md: no
    df_print: paged
  pdf_document:
    toc: yes
    df_print: kable
    number_sections: false
    fig_caption: yes
    highlight: tango
    dev: pdf
  word_document:
    toc: yes
    df_print: paged
    keep_md: no
---

## Course Project

**Regression Models Course Project**

Peer-graded Assignment

* This course project is available on GitHub

    <a href="https://github.com/oraclejavanet/regression-models-course-project" target="_top">Motor Trend Data Analysis</a>

## Synopsis

*Pending*

## Environment Setup

```{r setup, include = FALSE}
# set knitr options
knitr::opts_chunk$set(echo = TRUE, fig.path='figures/')

# clear global workspace (including hidden objects) for this session
rm(list = ls(all.names = TRUE))

# free up memory and display statistics on free memory
gc()

# disable scientific notation for numbers
options(scipen = 1)
```

Load packages used in this analysis.

```{r load-packages, echo = TRUE}
if (!require(ggplot2)) {
    install.packages("ggplot2")
    library(ggplot2)
}
if (!require("dplyr")) {
    install.packages("dplyr")
    library("dplyr")
}
```

Display session information.

```{r display-session-info, echo = TRUE}
sessionInfo()
```

## Load Data

Load and prepare the mtcars dataset. Convert categorical variables to factors.

```{r load-data, echo = TRUE}
library(datasets)
data(mtcars)
```

## Basic Data Summary

After loading the mtcars dataset, provide a basic summary of the data.

```{r basic-data-summary, echo = TRUE}
str(mtcars)
summary(mtcars)
head(mtcars)
dim(mtcars)

# tabulate delivery method and dosage level values
#table(ToothGrowth$supp, ToothGrowth$dose)

# summary of tooth length data grouped by delivery method and dosage level
#by(data = ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), summary)
```

## Exploratory Data Analysis

Perform some basic exploratory data analyses of the data. The analyses will
explore the following relations:

1. Relation 1
1. Relation 2

### Relation 1

```{r relation-1, echo = TRUE}
```

**Observation**

*Pending*

### Relation 2

```{r relation-2, echo = TRUE}
```

**Observation**

*Pending*



## Linear Models


*Pending*


## Analysis of Residuals


*Pending*


## Conclusion


*Pending*



